Networks Matter in International Relations:\\A Latent Network Approach to Dyadic Analysis

### 2017-10-27 paper/1_intro.tex
typos from MF
minor cleanup

In this article, we discuss the Additive and Multiplicative Effects (AME) model, 
a Bayesian approach for directly modeling relational data to reflect the data generating process that yields interdependencies in dyadic data structures \citep{hoff:2008,minhas:etal:2016:arxiv}. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads as there may be a particular actor that is more likely to send or receive events such as conflict. Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad; for example, if a rebel group initiates a conflict with a government, the government will likely reciprocate that behavior. We capture these two dependencies, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. The multiplicative effects capture dependencies that result from groups of actors clustering together or organizing into communities due to \textit{meso-scopic} features of networks, such as homophily and stochastic equivalence. These type of meso-scopic features often arise in relational data because actors possess some latent set of shared attributes that affect their probability of interacting with one another.

>>>>

In this article, we discuss the Additive and Multiplicative Effects (AME) model, 
a Bayesian approach for directly modeling relational data to reflect the data generating process that yields interdependencies in dyadic data structures \citep{hoff:2008,minhas:etal:2016:arxiv}. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads if a particular actor is more likely to send or receive events such as conflict. Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad; for example, if a rebel group initiates a conflict with a government, the government will likely reciprocate that behavior. We capture these two dependencies, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. The multiplicative effects capture dependencies that result from groups of actors clustering together or organizing into communities due to \textit{meso-scopic} features of networks, such as homophily and stochastic equivalence. These type of meso-scopic features often arise in relational data because actors possess some set of latent, shared attributes that affect their probability of interacting with one another.

### 2017-10-27 paper/4_app.tex
minor cleanup

We choose five prominent studies from the broad field of international relations and international political economy that utilize relational data \citep{reiter:stam:2003, mcdonald:2004,  rose:2004, weeks:2012, gibler:2017}. These studies are recent and have been cited over 100 times. Each of these pieces was published in a prominent journal and is well-known in the literature. Each used the standard approach in political science, which is to employ some form of a general linearized regression that ignores dyadic interdependencies, except as they may reveal themselves in included variables. In addition, each of these studies adjusts the posterior standard errors in an attempt to account for the clustering of observations. We follow the strategy of the original study in our re-estimation of the original models.

@@

One takeaway from is that many studies make knowledge claims based on differences in statistical significance between covariates. These differences may change dramatically when interdependencies are taken into account directly, as they do in this study.

@@

When we compare the results of Rose's original OLS model to the AME model accounting for network dependencies, the results are generally similar. The main result of the model---the null effect of membership in the WTO, as represented by the ``One-In" and ``Both-In" variables---does not prevail in the AME model. In the original, this negative linkages is none significant, but in the AME estimation a this negative linkage is more precisely estimated. However, while in the original model there was a clear positive relationship between Real GDP and Trade, most of this effect vanishes in the AME model. The random effects shown in Figure~\ref{fig:roser} reveal the cause of much of this divergence. Here, the states with the most positive random effects are also states with high GDP, though not necessarily high GDP/capita.\footnote{Note: Qatar exhibits strongly negative random effects.} Thus, the effect of GDP in the original model was, in part, an artifact of first-order dependencies. Most of the other results of the model are constant across each model, though some geographic features, such as islands and landlocked states, have a more clear effect on trade once we account for these network dependencies. When we account for network interdependencies, we observe a markedly lower Root Mean Squared Error out-of-sample --- $3.23$ for the OLS model and $1.77$ for the AME model. While the AME model replicate the main result, there are substantial and substantive improvements gleaned from the AME results.

@@

First and foremost, many findings which emerge from models that do not take interdependencies into account lose their statistical significance when network effects are estimated via AME.\footnote{This should not be news---since the finding has long been in the theoretical literature, but given the state of current literature in international relations is still pertinent.} Not only are coefficients biased in the GLM approaches to the analysis of dyadic data, but they are often imprecisely measured, with inflated standard errors.  This means that significant testing (for better or worse) is compromised when network effects are ignored.

>>>>

We choose five prominent studies from the broad field of international relations and international political economy that utilize relational data \citep{reiter:stam:2003, mcdonald:2004,  rose:2004, weeks:2012, gibler:2017}. These studies are recent and have been cited over 100 times. Each of these pieces was published in a prominent journal and is well-known in the literature. Each used the standard approach in political science, which is to employ some form of a general linearized regression that ignores dyadic interdependencies, except as they may reveal themselves in included variables. In addition, almost all of these studies adjusts the standard errors in an attempt to account for the clustering of observations. We follow the strategy of the original study in our re-estimation of the original models.

@@

One takeaway is that many studies make knowledge claims based on differences in statistical significance between covariates. These differences may change dramatically when interdependencies are taken into account directly, as they do in this study.

@@

When we compare the results of Rose's original OLS model to the AME model accounting for network dependencies, the results are generally similar. The main result of the model---the null effect of membership in the WTO, as represented by the ``One-In" and ``Both-In" variables---does not prevail in the AME model. In the original, this negative linkages is not significant, but in the AME estimation this negative linkage is more precisely estimated. However, while in the original model there was a clear positive relationship between Real GDP and Trade, most of this effect vanishes in the AME model. The random effects shown in Figure~\ref{fig:roser} reveal the cause of much of this divergence. Here, the states with the most positive random effects are also states with high GDP, though not necessarily high GDP/capita.\footnote{Note: Qatar exhibits strongly negative random effects.} Thus, the effect of GDP in the original model was, in part, an artifact of first-order dependencies. Most of the other results of the model are constant across each model, though some geographic features, such as islands and landlocked states, have a more clear effect on trade once we account for these network dependencies. When we account for network interdependencies, we observe a markedly lower Root Mean Squared Error out-of-sample --- $3.23$ for the OLS model and $1.77$ for the AME model. While the AME model replicate the main result, there are substantial and substantive improvements gleaned from the AME results.

@@

First and foremost, many findings which emerge from models that do not take interdependencies into account lose their statistical significance when network effects are estimated via AME.\footnote{This should not be news---since the finding has long been in the theoretical literature, but given the state of current literature in international relations is still pertinent.} Not only are coefficients biased in the GLM approaches to the analysis of dyadic data, but they are often imprecisely measured, with poorly calibrated standard errors.  This means that significant testing (for better or worse) is compromised when network effects are ignored.
